Type-Safe Modular Parsing

### 2017-01-14 paper/sections/ObjectAlgebras.tex
Section 4

The Expression Problem~\cite{wadler1998expression} illustrates the difficulty of extending data structures or ASTs in two dimensions. For traditional OO ASTs, adding new language constructs is easy, but adding new operations is hard. In the last section, we have a simple language containing literals, additions and variables. The corresponding ASTs are built using classes with a pretty-printing operation. Now suppose we want to collect free variables over the ASTs. By traditional OO approach, one attempt would be extending \inlinecode{Expr} with the new operation to obtain a new abstract type for ASTs:

@@

But then programmers have to define a new class for each construct by extending the old class with \lstinline{FVarsExpr}, and implementing the corresponding \lstinline{fVars} method:

@@

Besides introducing boilerplate code, more classes have to be defined along with future extensions. They pollutes the namespace heavily. More importantly, the \lstinline{VarExprParser} we defined is no longer suitable for parsing the new ASTs! All the methods in the previous parser has type \inlinecode{Parser[Expr]}, which indicates the result has type \inlinecode{Expr}, whereas the type should be \inlinecode{FVarsExpr} now.

Therefore, in order to generate new AST structures, we have to substitute \lstinline{FVarsExpr}, \lstinline{FVarsLit}, \lstinline{FVarsAdd} and \lstinline{FVarsVar} for the old \lstinline{Expr}, \lstinline{Lit}, \lstinline{Add} and \lstinline{Var} correspondingly. Again, to avoid modifying existing code, we have to sacrifice code reuse. In fact, there is a trade-off between separate compilation and type-safe code reuse. For semantic modularity, we need a different approach for building ASTs.

@@

Fortunately, Object Algebras~\cite{Oliveira:2012} enables us to solve this problem. As a nice solution, it separates data variants
and operations, and offers high flexibility in the choice of operations to be performed over ASTs. Before adopting it for parsing, the next section introduces the background of Object Algebras.

\subsection{Object Algebras}\label{subsec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem, achieving two dimensions of extensibility (language constructs and operations) in a modular and type-safe way. The definition of data structures is separated from their behaviours, and future extensions on both dimensions no longer require existing code to be modified, supporting separate compilation.

@@

\subsection{Discussion of Object Algebras}\label{subsec:more-oa}

In Section~\ref{subsec:objectalgebras}, \lstinline{makeExp} represents the AST of expression \inlinecode{1 + x} as a method. Parsers with Object Algebras generate such ``implicit objects'', which are actually functions of type \lstinline{Alg[E] => } \lstinline{E} for generic \lstinline{E}. Comparing with traditional objects, one may have concern that if a parser produces function structures, it could be hard to further process them. Especially, in the design of a compiler there will be a lot of desugarings (or AST transformations).

However, we argue that this is not an issue, because functions can be used
in a very clever way. Continuing with the previous example, suppose we want to transform an expression of \inlinecode{VarExprAlg} to an expression of \inlinecode{ExprAlg}, by
replacing some variables with values based on a variable environment. A \textit{transformation} algebra can be implemented as follows:

@@

Another useful technique is the composition of \textit{query} algebras. A query traverses the abstract syntax tree, collects information and returns the result. The aforementioned \lstinline{Print} and \lstinline{Eval} algebras are traditional queries. But as we discussed before, an algebra is first fed to the parser, then \lstinline{parse} is invoked to do the parsing. It is tedious to fed \lstinline{Print} and \lstinline{Eval} separately, as the same input is parsed twice. Instead, we can firstly compose two queries into one~\cite{Oliveira:2012,oliveira2013feature}, before applying it for parsing.

>>>>

The Expression Problem~\cite{wadler1998expression} illustrates the
difficulty of extending data structures or ASTs in two dimensions. For
traditional OO ASTs, adding new language constructs is easy, but
adding new operations is hard. In the last section, we have a simple
language containing literals, additions and variables. The
corresponding ASTs are built using classes with a pretty-printing
operation. Now suppose we want to collect free variables over the
ASTs. Using a traditional OO approach, one attempt would be extending
\inlinecode{Expr} with the new operation to obtain a new abstract type
for ASTs:

@@

\noindent Then all classes represented language constructs could be 
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work 
by Wang and Oliveira~\cite{wang2016expression}, which show a technique that is type-safe
in many cases). More importantly, a second problem is that
\begin{comment}
But then programmers have to define a new class for each construct by
extending the old class with \lstinline{FVarsExpr}, and implementing
the corresponding \lstinline{fVars} method:

@@

Besides introducing boilerplate code, more classes have to be defined
along with future extensions. They pollutes the namespace heavily.
More importantly, 
\end{comment}
is that even if that approach would work, the parsing code would not
be reusable. \lstinline{VarExprParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser has type \inlinecode{Parser[Expr]}, which indicates the result have type \inlinecode{Expr}, whereas the type should be \inlinecode{FVarsExpr} now.

Therefore, in order to generate new AST structures, we have to
substitute \lstinline{FVarsExpr}, \lstinline{FVarsLit},
\lstinline{FVarsAdd} and \lstinline{FVarsVar} for the old
\lstinline{Expr}, \lstinline{Lit}, \lstinline{Add} and \lstinline{Var}
correspondingly. Thus we have to either modify existing code or 
use casts. For semantic modularity, we need a different approach for building ASTs.

@@

Fortunately, Object Algebras~\cite{Oliveira:2012} enable us to solve
this problem. Object Algebras separate data variants and
operations, and offer high flexibility in the choice of operations to
be performed over ASTs. The next section provides a brief introduction 
to Object Algebras.

\subsection{Object Algebras}\label{subsec:objectalgebras}

Object Algebras captures a design pattern that address the Expression
Problem, achieving two dimensions of extensibility (language
constructs and operations) in a modular and type-safe way. The
definition of data structures is separated from their behaviours, and
future extensions on both dimensions no longer require existing code
to be modified, supporting separate compilation.

@@

\begin{comment}
\subsection{ASTs and Object Algebras}\label{subsec:more-oa}

In Section~\ref{subsec:objectalgebras}, \lstinline{makeExp} represents
the AST of expression \inlinecode{1 + x} as a method. Parsers with
Object Algebras generate such ``implicit objects'', which are actually
functions of type \lstinline{Alg[E] => } \lstinline{E} for generic
\lstinline{E}. Comparing with traditional objects, one may have worry
that if a parser produces function structures, it could be hard to
further process them. Especially, in the design of a compiler there
will be a lot of desugarings (or AST transformations). As previous
work on Object Alegbras has shown~\cite{Gouseti2014,Zhang2015}, AST
transformations can be implemented using \emph{transformation
  algebras}. Continuing with the previous example, suppose we want to
transform an expression of \inlinecode{VarExprAlg} to an expression of
\inlinecode{ExprAlg}, by replacing some variables with values based on
a variable environment. A \textit{transformation} algebra is
implemented as follows:

@@

Another useful technique is the composition of \textit{query} algebras. A query traverses the abstract syntax tree, collects information and returns the result. The aforementioned \lstinline{Print} and \lstinline{Eval} algebras are traditional queries. But as we discussed before, an algebra is first fed to the parser, then \lstinline{parse} is invoked to do the parsing. It is tedious to fed \lstinline{Print} and \lstinline{Eval} separately, as the same input is parsed twice. Instead, we can firstly compose two queries into one~\cite{Oliveira:2012,oliveira2013feature}, before applying it for parsing.
\end{comment}
